Model-agnostic Interpretable Machine Learning

### 2021-08-16 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
continue draft

---
title: "Model-agnostic Interpretable Machine Learning"
author: "Christoph Molnar"
date: "2021-08"
institute: "BIPS bremen"
output: 
  beamer_presentation:
    includes:
      in_header: "header_pagenrs.tex"
---

@@

# Some ideas

- Cite pitfalls at many places in the talk
- INtroduction with History: INterweave motivation, and also cite history paper
-

# Outline

- A Brief History of Interpretable Modeling
- A Closer Look at Model-agnostic interpretability
- methods:
  - Feature Effects: PDP, ICE and ALE
  - Feature Importance: PFI
- Overview of Stuff I did in my PhD 
- Measuring Model Complexity
- Conditional Feature Importance Using Subgroups
- Inference of Effects

# Before You Pin Me Down

@@

# A Brief History of Interpretable Modeling

Interpretable Models with data have a longer history, but the field of interpretable machine learning is still very young as consolidated field.

* 1800s: Linear regression models
* 1960: Rule-based ML (AID, CHAID, CART, association rules, ...)
* 2001: Random Forest
* 2021: Deep Learning Boom kicked off
* 2016: Interpretable ML / Explainable AI boom

While 200 years of history, consolidation as larger field only started in 2016.

@@

# Why Interpretability?

@@

# Interpretable Models

@@

@@

# PDP and ICE

# PFI

TODO: Explain how PFI works

# PFI and PDP similarity

@@

## Soluation: Conditional variants

@@

## Thank You for Your Attention

>>>>

---
title: "Model-agnostic Interpretable Machine Learning"
author: "Christoph Molnar"
date: "2021-08"
institute: "BIPS bremen"
output: ioslides_presentation
---

@@

## Some ideas

- Cite pitfalls at many places in the talk
- INtroduction with History: INterweave motivation, and also cite history paper

## Outline

- A Brief History of Interpretable Modeling
- Model-agnostic interpretability
- Methods PDP and PFI
- What I did in my PhD 
    - Conditional Feature for PFI and PDP Importance Using Subgroups
    - Statistical Inference for PDP and PFI
    - Measuring Model Complexity

## Before You Pin Me Down

@@

## A Brief History of Interpretable Modeling

* 1800s: Linear regression models
* 1960s: Rule-based ML
* 2001:  Random Forest
* 2012:  Deep Learning Boom kicked off
* 2016:  Interpretable ML / Explainable AI boom

Interpretable ML has 200 year old roots, but consolidation only started in 2016.

@@

## Why Interpretability?

@@

## Interpretable Models

@@

## Interpretable Models

```{r, echo = FALSE}
knitr::include_graphics("../images/white-box.png")
```

@@

## Interpreting Models

```{r, echo=FALSE}
knitr::include_graphics("../images/black-box.png")
```

## Model-agnostic Methods

```{r, echo=FALSE}
knitr::include_graphics("../images/agnostic-black-box.png")
```

# Methodological Background

## Individual Conditional Expectations (ICE)

maybe use ice + pdp animation, but split into individual frames so they are clickable

## Partial Dependence Plot (PDP)

The theoretical construct:
$$ PD(x) = \mathbb{E}_{X_{C}}[\hat{f}(x, X_{C})] $$

The estimator:
$$\widehat{PD} = \frac{1}{n_2}\sum_{i=1}^{n_2} \hat{f}(x, x_{C}^{(i)})$$

What it looks like:

TODO: 1D-plot and 2D-plot

## Permutation Feature Importance PFI

The construct:
$$PFI =  \mathbb{E}_{Y \tilde{X}_S X_C}(L(Y, \hat{f}(\tilde{X}_S, X_C))) - \mathbb{E}_{Y X_S X_C}(L(Y,\hat{f}(X_S, X_C)))$$

The algorithm (simples form): CITE Breiman

- Measure original loss $L$ on test data.
- Permute features $X_S$ to get sample $\tilde{X}_S$.
- Get prediction with $\tilde{X}_S$, measure loss again.
- Measure difference between losses.
- Interpretation: PFI is performance drop when "destroying" information about feature.

## PFI and PDP similarity

@@

## Solution: Conditional variants

@@

# Inference of Feature Effects and Importance in Machine Learning

## Status-Quo

- For inference, practitioner rely on interpretable models
- remember: interpretable models have parameters that map to data-generating process
- but: situations where more flexible models match DGP better (measured in test loss)
- we have: model-agnostic interpretation
- But: only describes the model behavior
- Question: Can we link PD and PFI to the DGP? and even do inference?

## Example

TODO: Show PDP and PFI from some papers?

## Linking external tools to DGP

- Hope: Lift PDP and PFI to status of statistical estimator
- Needed: estimand. A definition of what it is supposed to estimate in the DGP (not the model)
- Idea: PDP and PFI are expectation of a function
- Assumption: True function $f$ exists in DGP
- Then: We can apply same interpretation tools to DGP. Very simple.

TODO: simple image

## Source of uncertainty

- The ground truth estimand allows use to do the usual statistical estimator stuff:
- For example: Measure MSE
- Or: split up MSE into bias (systematic deviation of estimator to estimand)
- Variance: Two sources in the estimation process:
  - Model is a random variable bc. training data is a sample + model-specific inherent randomness (such as subset sampling in random forests)
  - Expectations are based on Monte carlo sampling

## Model-PD and Model-PFI

Show definition plus confidence intervals plus image

## Learner-PD and Learner-PFI

Show definition plus confidence intervals plus image

But there is a problem when estimating this variance

## Variance correction

- Variance is underestimated as training data is shared by models.
- Same for test data.
- Underestimating variance also means confidence intervals are too short, appearing more certain than they really are.
- Same problem exists for variance estimates for model performance, as needed for comparing models
- Nadeau et. al (CITE) proposed correction term
- term very naive, but MUCH better than without correction

## Show simulation results

# Thank You for Your Attention
